draft

### 2025-05-25 draft.tex
Updates from Overleaf

\begin{abstract}
Biological image analysis has traditionally focused on measuring specific visual properties of interest for cells or other entities. A complementary paradigm gaining increasing traction is image-based profiling - quantifying many distinct visual features to form comprehensive profiles where patterns of changes show biological properties. While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering ML workflows. Here we introduce cp\_measure, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction. We demonstrate that cp\_measure features retain high fidelity with CellProfiler measurements while enabling seamless integration with the scientific Python ecosystem. Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how cp\_measure enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
\end{abstract}

\section{Introduction}
\label{sec:org4c9ba67}
Modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components by microscopy.

Historically, the human eyes and brain drew conclusions from biological images but this approach is limited in scale, objectivity, and reproducibility. Now, quantification is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.
We use the terms "features" and "measurements" interchangeably here.

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object"). It can featurize the pairwise combination of all the available channels (colours) and objects. The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
\end{figure}

\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. It can be used by experimental biologists with scant programming experience, for whom it has been a boon. By contrast, for computer-savvy scientists, its graphical user interface and lack of programmatic access means friction in pipelines comprised of multiple tools. CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance. For example, mix and matching CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces. The usage of CellProfiler also precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction. Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats.

Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} . However, these have limitations: First, their implementation is completely independent from CellProfiler and thus they do not produce the same feature set as CellProfiler, precluding matching to existing datasets such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}. Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset. Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.  Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts \cite{garcia-fossaInterpretingImagebasedProfiles2023}.

In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract engineered features while readily integrating with computational workflows to achieve reproducible data analyses with a few lines of code, whether locally or in the cloud. We thus believe that there is the need for a readily-integrated feature extraction tool aimed at generalist and high throughput problems not being solved by the current software ecosystem. With all this in mind, we developed cp\_measure.

\section{Extracting interpretable features}
\label{sec:org61842b5}
The library cp\_measure is  the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task. Measurements are categorized based on three kinds of input: One object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Pearson correlations between intensity values across channels) and multiple objects (e.g., number of neighbours).

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogs. \textbf{Left panel.} Representative examples comparing CellProfiler feature values to cp\_measure's, generated using matching pairs of masks and images. \textbf{Right panel.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
\end{figure}

@@

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features. The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover with relatively few lines of code.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these features as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological features yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figs/spatial.png}
\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
\end{figure}
\section{Discussion}
\label{sec:orgf37b369}
While point-and-click interfaces open the world of image analysis to many researchers, they can present barriers for smooth adoption into computational workflows. In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks). Although there are still rough edges to polish, such as ensuring features match 1:1, it enables simpler automated profiling of microscopy data in short scripts and complex pipelines. It avoids graphical interfaces, resulting in better scaling capabilities for high-content microscopy, with or without cloud infrastructure.

The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology. When used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.

Already, Cellprofiler features have been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. By association, we foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
\section{Future work}
\label{sec:org5cdbb12}
We propose in future work to contribute cp\_measure back to CellProfiler, which offers several benefits. It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in the leaner and more approachable cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.

We also plan to develop a comprehensive test suite to guarantee mathematical correctness, which currently CellProfiler itself is lacking. A functioning test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity. Once tests are in place, we could add support for just-in-time compiling and GPUs. We envision the cp\_measure library could also be the place to develop and distribute new measurements from the community.

>>>>

\begin{abstract}
Biological image analysis has traditionally focused on measuring specific visual properties of interest for cells or other entities. A complementary paradigm gaining increasing traction is image-based profiling - quantifying many distinct visual features to form comprehensive profiles where patterns of changes show biological properties. While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering ML workflows. Here we introduce cp\_measure, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction. We demonstrate that cp\_measure features retain high fidelity with CellProfiler features while enabling seamless integration with the scientific Python ecosystem. Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how cp\_measure enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
\end{abstract}

\section{Introduction}
\label{sec:org4c9ba67}
High throughput screening of biological phenomena via complex modalities, such as RNA sequencing, is prohibitively expensive, therefore microscopy is an efficient first step. Through microscopy, modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components. Nowadays, quantification of biological images is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.99\linewidth]{./figs/fig_1_tmp.png}
\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object"). It can featurize the pairwise combination of all the available channels (colours) and objects. The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
\end{figure}

\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. It can be used by experimental biologists with scant programming experience, for whom it has been a boon. By contrast, it may not suit the needs of computer-savvy scientists building high throughput pipelines comprised of multiple tools. CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance. For example, complementing CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces. Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts. Containers mitigate this problem, but not without increased complexity and caveats.

Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} . However, these come with constraints: Given that their implementation is completely independent from CellProfiler, they do not produce the same feature set, precluding matching to existing datasets, such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}. Moreover, a common problem is that these tools enforce a structure for the input data, such as determined hierarchical file structures, which require significant wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a pretrained deep-learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Compared to engineered features, deep learning approaches may pose a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset, in addition to having access to a GPU. Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.  Unlike learned features, whose interpretation requires complex methods like counterfactual techniques or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts, such as the colocalisation of two proteins \cite{garcia-fossaInterpretingImagebasedProfiles2023}.

In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract interpretable features while readily integrating with computational workflows to achieve reproducible data analyses, whether locally or in the cloud. We thus believe that there is the need for a readily-integrated feature extraction tool aimed at generalist and high throughput problems not being solved by the current software ecosystem. With all this in mind, we developed cp\_measure.

\section{Extracting interpretable features}
\label{sec:org61842b5}
The library cp\_measure branches off the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task. Measurements are defined as a collection of related features, following CellProfiler's nomenclature, and are categorized based on three kinds of input: One object (producing features such as area and eccentricity), one object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Manders correlations between intensity values across channels).

@@

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogs. \textbf{Left panel.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images. \textbf{Right panel.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
\end{figure}

@@

Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features. The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.
\subsubsection{Spatial transcriptomics}
\label{sec:org5711d86}
A key advantage of providing these features as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment. The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality. However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment. In contrast, classical morphological features yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figs/spatial.png}
\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
\end{figure}
\section{Discussion}
\label{sec:orgf37b369}
While point-and-click interfaces open the world of image analysis to many researchers, they are not as effective for computational workflows with no human-in-the-loop. In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks). It enables simpler automated profiling of microscopy data in short scripts and complex pipelines. The modularity it provides facilitates pipelines with better scaling capabilities for high-content microscopy, with or without cloud infrastructure.

The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology. A potential workflow would be to use deep learning features to cluster and then use cp\_measure features from these clusters. As a whole, when used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.

Already, Cellprofiler features have been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}. By association, we foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
\section{Future work}
\label{sec:org5cdbb12}
We propose in future work to make it possible for CellProfiler to import cp\_measure, which offers several benefits. It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements. It would also mean that any changes made in cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.

We also plan to develop a comprehensive test suite to guarantee mathematical correctness, which currently CellProfiler itself is lacking. A comprehensive test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity, by providing rapid iteration capabilities. Once tests are in place, we could add support for just-in-time compiling and GPUs. We envision that cp\_measure could also be the place to develop and distribute new measurements from the community.
